Adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2024-03-07 ms.tex

For our point-particle source on a circular geodesic, the radial mode function $R_{-2jm}(r)$ have the asymptotic form
\begin{subequations} \label{eqn:asympR}
\begin{align}
    R_{-2jm}(r\rightarrow r_+) &\simeq Z^\mathcal{H}_{-2jm} \Delta^{2} e^{-im k_p r_*},
    \\ 
    R_{-2jm}(r\rightarrow \infty) &\simeq Z^\mathcal{I}_{-2jm} r^{3} e^{+im\Omega_p r_*},
\end{align}
\end{subequations}
where $k_p = \Omega_p - \Omega_+$, the horizon frequency $\Omega_+ = a/(2Mr_+)$, and the tortoise coordinate $r_*$ is given by the differential relation $dr_*/dr = (r^2+a^2)/\Delta$.
The amplitudes $Z^\mathcal{H/I}_{-2jm}$ are often referred to as \emph{Teukolsky amplitudes} and are constructed via the standard Green's function method, also known as the method of variation of parameters (see Sec.~III\,A in \cite{HughETC21} or \cite{DrasHugh06} for further details). From \eqref{eqn:XfromR}, we see that $X_{-2\ell m}$ then possesses the same asymptotic behavior as $R_{-2\ell m}$ in \eqref{eqn:asympR} but with modified amplitudes
\begin{align}
    Z^\mathcal{H/I}_{-2jm} \rightarrow X^\mathcal{H/I}_{-2\ell m} = \sum_{j=\ell_\mathrm{min}}^\infty b^\ell_{-2jm}(\gamma) Z^\mathcal{H/I}_{-2jm}.
\end{align}
As a final note, because $Z^\mathrm{H/I}_{-2jm}$ and  $X^\mathrm{H/I}_{-2\ell m}$ depend on the source motion, we can parametrize both amplitudes in terms of the orbital constants, e.g., $X^\mathcal{H/I}_{-2\ell m}= {X}^\mathcal{H/I}_{-2\ell m}(\Omega_p; a)$.

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/timeAndPhaseConvergence.pdf}
    \caption{Convergence of the maximum absolute interpolation error for the frequency-weighted time and phase splines, $||\Delta^{\Omega t}_{(n,n)}||_\infty$ (orange squares) and $||\Delta^\Phi_{(n,n)}||_\infty$ (green diamonds), along with the fractional interpolation error for the unweighted time-spline $||\Delta^t_{(n,n)}||_\infty$ (blue diamonds). The dashed lines correspond to a power-law convergence of $2^{4n}$.}
    \label{fig:trajConvergence}
\end{figure}

@@

As a final check, we perform a parallel-tempered Markov Chain Monte Carlo (MCMC) sampling of the posterior distribution for intrinsic the source parameters $(M, \mu, \hat{a}, p_0, \Phi_{\phi0}) = (10^6 M_\odot, 10^5 M_\odot, 0.9, 12.55, 0.2)$ using the open-source \texttt{Eryn} sampling tool \cite{ForeETC13, KarnETC23,Katz23}.\footnote{Note that we choose a very comparable mass ratio to speed-up waveform evaluation and the calculation of the likelihood. This allows us to perform the sampling within a day on a laptop. More realistic EMRI systems are much more computationally expensive to sample and beyond the scope of this work. MCMC sampling of astrophysically-realistic EMRIs with \texttt{bhpwave} will be presented in a following paper.} To simplify the calculation, we only sample over the intrinsic parameters $(\log M, \log \mu, a, p_0, \Phi_{\phi0})$ and hold all other parameters fixed. We then perform our Fisher analysis on the same system. In Figure \ref{fig:mcmcOverlay} we plot the two dimensional contours of the posterior distribution calculated from our MCMC sampling (dark blue lines). Each contour line corresponds to a one-$\sigma$ deviation from a neighboring contour. We then use the covariances predicted from our Fisher analysis to sample a multivariate distribution and overlay this on top of our MCMC results (lighter orange lines). As we can see, the two sets of samples lie nearly on top of one another, indicating that our Fisher matrix analysis provides a leading-order estimate for the shape of the posterior.

>>>>

For our point-particle source on a circular geodesic, the radial mode function $R_{-2jm}(r)$ have the asymptotic form
\begin{subequations} \label{eqn:asympR}
\begin{align}
    R_{-2jm}(r\rightarrow r_+) &\simeq Z^\mathcal{H}_{-2jm} \Delta^{2} e^{-im k_p r_*},
    \\ 
    R_{-2jm}(r\rightarrow \infty) &\simeq Z^\mathcal{I}_{-2jm} r^{3} e^{+im\Omega_p r_*},
\end{align}
\end{subequations}
where $k_p = \Omega_p - \Omega_+$, the horizon frequency $\Omega_+ = a/(2Mr_+)$, and the tortoise coordinate $r_*$ is given by the differential relation $dr_*/dr = (r^2+a^2)/\Delta$.
The amplitudes $Z^\mathcal{H/I}_{-2jm}$ are often referred to as \emph{Teukolsky amplitudes} and are constructed via the standard Green's function method, also known as the method of variation of parameters (see Sec.~III\,A in \cite{HughETC21} or \cite{DrasHugh06} for further details). From \eqref{eqn:XfromR}, we see that $X_{-2\ell m}$ then possesses the same asymptotic behavior as $R_{-2\ell m}$ in \eqref{eqn:asympR} but with modified amplitudes
\begin{align}
    Z^\mathcal{H/I}_{-2jm} \rightarrow X^\mathcal{H/I}_{-2\ell m} = \sum_{j=\ell_\mathrm{min}}^\infty b^\ell_{-2jm}(\gamma) Z^\mathcal{H/I}_{-2jm}.
\end{align}
As a final note, because $Z^\mathcal{H/I}_{-2jm}$ and  $X^\mathcal{H/I}_{-2\ell m}$ depend on the source motion, we can parametrize both amplitudes in terms of the orbital constants, e.g., $X^\mathcal{H/I}_{-2\ell m}= {X}^\mathcal{H/I}_{-2\ell m}(\Omega_p; a)$.

@@

\begin{figure}[!t]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/timeAndPhaseConvergence.pdf}
    \caption{Convergence of the maximum absolute interpolation error for the frequency-weighted time and phase splines, $||\Delta^{\Omega t}_{(n,n)}||_\infty$ (orange squares) and $||\Delta^\Phi_{(n,n)}||_\infty$ (green diamonds), along with the fractional interpolation error for the unweighted time-spline $||\Delta^t_{(n,n)}||_\infty$ (blue diamonds). The dashed lines correspond to a power-law convergence of $2^{-4n}$.}
    \label{fig:trajConvergence}
\end{figure}

@@

As a final check, we perform a parallel-tempered Markov Chain Monte Carlo (MCMC) sampling of the posterior distribution for intrinsic the source parameters $(M, \mu, \hat{a}, p_0, \Phi_{\phi0}) = (10^6 M_\odot, 10^5 M_\odot, 0.9, 12.55, 0.2)$ using the open-source \texttt{Eryn} sampling tool \cite{ForeETC13, KarnETC23,KatzKarnKors23}.\footnote{Note that we choose a very comparable mass ratio to speed-up waveform evaluation and the calculation of the likelihood. This allows us to perform the sampling within a day on a laptop. More realistic EMRI systems are much more computationally expensive to sample and beyond the scope of this work. MCMC sampling of astrophysically-realistic EMRIs with \texttt{bhpwave} will be presented in a following paper.} To simplify the calculation, we only sample over the intrinsic parameters $(\log M, \log \mu, a, p_0, \Phi_{\phi0})$ and hold all other parameters fixed. We then perform our Fisher analysis on the same system. In Figure \ref{fig:mcmcOverlay} we plot the two dimensional contours of the posterior distribution calculated from our MCMC sampling (dark blue lines). Each contour line corresponds to a one-$\sigma$ deviation from a neighboring contour. We then use the covariances predicted from our Fisher analysis to sample a multivariate distribution and overlay this on top of our MCMC results (lighter orange lines). As we can see, the two sets of samples lie nearly on top of one another, indicating that our Fisher matrix analysis provides a leading-order estimate for the shape of the posterior.
